Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-22 prop_letter.tex
prop_letter.tex \section{Born rule and Gleason's Theorem} Updating...

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
there is always a state consistent with the composited QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$~\citep{HOSTunpublished}.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. To state this result, we introduce a few technical notions.
First, the norm of a pure state~$\ket{\psi}$ is defined as usual
$\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$. Then, the norm
of the difference between two states~$\rho$ and $\rho'$ is just
the usual operator norm
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,,
\end{equation}
which is also commonly known as $2$-norm or spectral norm~\citep{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.

\begin{thm}[Finite-Precision Extension of Weak Gleason's theorem]\label{thm:Finite-precision-Gleason}Consider
an interval-valued function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
with the maximum length of intervals $\delta$. Given any state~$\rho$,
if a state $\rho'\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then the norm of their difference is $\left\Vert \rho-\rho'\right\Vert \le\delta$.\end{thm}

@@

Since $\iota$ maps to intervals, we can decompose it into two functions:
its left-end and right-end $\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
By Eq.~(\ref{eq:iota-Inclusion}), the left-end function $\iotal:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-convex~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992},
i.e., 
\begin{equation}
\begin{aligned} & \iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\\
\ge{} & \iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)\,.
\end{aligned}
\end{equation}
Then, for any $x_{0}'$ and $x_{1}'\in\left[0,1\right]$ with $x_{0}'\ge x_{1}'$,
we can substitute $x_{0}$ and $x_{1}$ by $\left(x_{0}'-x_{1}'\right)/2$
, substitute $x_{2}$ by $x_{1}'$, and get the following inequality
\begin{eqnarray}
\frac{\iotal\left(x_{0}'\right)+\iotal\left(x_{1}'\right)}{2} & \ge & \iotal\left(\frac{x_{0}'+x_{1}'}{2}\right)\,,
\end{eqnarray}
i.e., $\iotal$ is midpoint-convex or Jensen-convex~\citep{Klee_1956,Kenyon_1956,RobertsVarberg1973,PecaricTong1992}.
Together with the fact that $\iotal$ maps to a bounded interval~$\left[0,1\right]$,
the left-end function~$\iotal$ must be continuous on the unit open
interval~$\left(0,1\right)$~\citep{Jensen_1906,RobertsVarberg1973}.
Therefore, either $\iota$ maps every number in $\left(0,1\right)$
to the same interval, or the number of intervals that $\iota$ maps
to must be uncountable.

Even if $\iota$ has to be a THOS rule to guarantee $\iota\circ\mu$
to be a QIVPM, the composited function~$\iota\circ\mu$ may still
be a QIVPM when $\iota$ is not a THOS rule. In this case, the composited
function~$\iota\circ\mu$ could only map to a finite set of intervals.
For example, we can weaken the convex condition Eq.~(\ref{eq:iota-Inclusion})
to the following additive condition: for any numbers~$x_{0}$ and
$x_{1}\in\left[0,1\right]$, their sum~$x_{0}+x_{1}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, given a Hilbert space of dimension~$d\le3$, when $\iota$
satisfies Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add}),
$\iota\circ\mu$ still must be a QIVPM~\citep{HOSTunpublished}.
More concretely, given a three-dimensional Hilbert space, consider
two states~$\rho_{0}=\frac{1}{3}\cdot\proj{0}+\frac{1}{3}\cdot\proj{1}+\frac{1}{3}\cdot\proj{2}$
and $\rho_{1}=\frac{7}{24}\cdot\proj{0}+\frac{1}{3}\cdot\proj{1}+\frac{3}{8}\cdot\proj{2}$,
a set of intervals
\begin{equation}
\begin{aligned}\mathscr{I}_{2}={} & \left\{ \imposs,\left[0,\tfrac{1}{4}\right],\left[0,{\textstyle \tfrac{1}{2}}\right],\left[\tfrac{1}{4},\tfrac{3}{4}\right],\right.\\
 & \left.\left[\tfrac{1}{2},1\right],\left[\tfrac{1}{4},1\right],\necess\right\} 
\end{aligned}
\end{equation}
and a interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$
defined in Table~\ref{tab:I_2}.
\begin{table}
\caption{\label{tab:I_2}A interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$.}
\centering{}%
\begin{tabular}{cc}
\toprule 
\addlinespace
$x$ & $\iota_{2}\left(x\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$x=0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0<x\le\tfrac{1}{8}$ & $\left[0,\tfrac{1}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{1}{8}<x<\tfrac{3}{8}$ & $\left[0,{\textstyle \tfrac{1}{2}}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{3}{8}\le x\le\tfrac{5}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{5}{8}<x<\tfrac{7}{8}$ & $\left[\tfrac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{7}{8}\le x<1$ & $\left[\tfrac{1}{4},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$x=1$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Since $\iota_{2}$ satisfies Eqs.~(\ref{eq:iota-constraints}) and
Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}, $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$ are both QIVPMs.
By Thm.~\ref{thm:Finite-precision-Gleason}, $\rho_{0}\in\coreBorn\left(\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}},\events\right)$
and $\rho_{1}\in\coreBorn\left(\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$
implies $\left\Vert \rho_{0}-\rho_{1}\right\Vert \le\frac{1}{2}$.
This is true because the exact value of $\left\Vert \rho_{0}-\rho_{1}\right\Vert $
is $\frac{1}{24}$.

>>>>

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
there is always a state consistent with the composited QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$~\citep{HOSTunpublished}.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. Eventually, the collection of consistent states collapses
to the singleton set~$\left\{ \rho\right\} $ as the intervals becomes
the infinite precise intervals~$\mathscr{I}_{\infty}$. This fact
is essentially Gleason's theorem, but a little bit weaker because
we only consider the composited QIVPMs~$\iota\circ\mu_{\rho}^{\mathrm{B}}$
instead of general QIVPMs~$\bar{\mu}$. To state this result, we
introduce a few technical notions. First, the norm of a pure state~$\ket{\psi}$
is defined as usual $\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$.
Then, the norm of the difference between two states~$\rho$ and $\rho'$
is just the usual operator norm
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,,
\end{equation}
which is also known as $2$-norm or spectral norm~\citep{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.

\begin{thm}[Finite-Precision Extension of Weak Gleason's theorem]\label{thm:Finite-precision-Gleason}Consider
an interval-valued function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
with the maximum length of intervals $\delta$. Given any state~$\rho$,
if a state $\rho'\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then the norm of their difference $\left\Vert \rho-\rho'\right\Vert \le\delta$.\end{thm}

@@

Since $\iota$ maps to intervals, we can decompose it into two functions:
its left-end and right-end $\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
By Eq.~(\ref{eq:iota-Inclusion}), the left-end function $\iotal:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-convex~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992},
i.e., 
\begin{equation}
\begin{aligned} & \iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\\
\ge{} & \iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)\,.
\end{aligned}
\end{equation}
Then, for any $x_{0}'$ and $x_{1}'\in\left[0,1\right]$ with $x_{0}'\ge x_{1}'$,
we can substitute $x_{0}$ and $x_{1}$ by $\left(x_{0}'-x_{1}'\right)/2$
, substitute $x_{2}$ by $x_{1}'$, and get the following inequality
\begin{eqnarray}
\frac{\iotal\left(x_{0}'\right)+\iotal\left(x_{1}'\right)}{2} & \ge & \iotal\left(\frac{x_{0}'+x_{1}'}{2}\right)\,,
\end{eqnarray}
i.e., $\iotal$ is midpoint-convex or Jensen-convex~\citep{Klee_1956,Kenyon_1956,RobertsVarberg1973,PecaricTong1992}.
Together with the fact that $\iotal$ maps to a bounded interval $\left[0,1\right]$,
the left-end function~$\iotal$ must be continuous on the unit open
interval $\left(0,1\right)$~\citep{Jensen_1906,RobertsVarberg1973}.
Therefore, either $\iota$ maps every number in $\left(0,1\right)$
to the same interval, or the number of intervals that $\iota$ maps
to must be uncountable.

Even if $\iota$ has to be a THOS rule to guarantee $\iota\circ\mu$
a QIVPM, the composited function $\iota\circ\mu$ may still be a QIVPM
when $\iota$ is not a THOS rule. In this case, the composited function
$\iota\circ\mu$ could only map to a finite set of intervals. For
example, we can weaken the convex condition Eq.~(\ref{eq:iota-Inclusion})
to the following additive condition: for any numbers $x_{0}$ and
$x_{1}\in\left[0,1\right]$, their sum $x_{0}+x_{1}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, given a Hilbert space of dimension~$d\le3$, when $\iota$
satisfies Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add}),
$\iota\circ\mu$ still must be a QIVPM~\citep{HOSTunpublished}.
More concretely, given a three-dimensional Hilbert space, consider
two states~$\rho_{0}=\frac{1}{3}\cdot\proj{0}+\frac{1}{3}\cdot\proj{1}+\frac{1}{3}\cdot\proj{2}$
and $\rho_{1}=\frac{7}{24}\cdot\proj{0}+\frac{1}{3}\cdot\proj{1}+\frac{3}{8}\cdot\proj{2}$,
a set of intervals
\begin{equation}
\begin{aligned}\mathscr{I}_{2}={} & \left\{ \imposs,\left[0,\tfrac{1}{4}\right],\left[0,{\textstyle \tfrac{1}{2}}\right],\left[\tfrac{1}{4},\tfrac{3}{4}\right],\right.\\
 & \left.\left[\tfrac{1}{2},1\right],\left[\tfrac{1}{4},1\right],\necess\right\} 
\end{aligned}
\end{equation}
and a interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$
defined in Table~\ref{tab:I_2}.
\begin{table}
\caption{\label{tab:I_2}A interval-valued function~$\iota_{2}:\left[0,1\right]\rightarrow\mathscr{I}_{2}$.}
\centering{}%
\begin{tabular}{cc}
\toprule 
\addlinespace
$x$ & $\iota_{2}\left(x\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$x=0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0<x\le\tfrac{1}{8}$ & $\left[0,\tfrac{1}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{1}{8}<x<\tfrac{3}{8}$ & $\left[0,{\textstyle \tfrac{1}{2}}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{3}{8}\le x\le\tfrac{5}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{5}{8}<x<\tfrac{7}{8}$ & $\left[\tfrac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\tfrac{7}{8}\le x<1$ & $\left[\tfrac{1}{4},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$x=1$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Since $\iota_{2}$ satisfies Eqs.~(\ref{eq:iota-constraints}) and
Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}, $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$ are both QIVPMs.
By Thm.~\ref{thm:Finite-precision-Gleason}, $\rho_{0}\in\coreBorn\left(\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}},\events\right)$
and $\rho_{1}\in\coreBorn\left(\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$
implies $\left\Vert \rho_{0}-\rho_{1}\right\Vert \le\frac{1}{2}$.
This is true because the exact value of $\left\Vert \rho_{0}-\rho_{1}\right\Vert $
is $\frac{1}{24}$.
